plyranges: a grammar for transforming genomics data

### 2018-04-16 paper.Rmd
design section redraft

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogenous genomics data heterogenous genomics data by expresssing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED.An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use an 
SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric 
Query Language (GMQL) implements a relational algebra for combining genomic datasets 
@Kaitoua2017-pw.  The command line application BEDtools develops an extensive 
algebra for performing arithmetic between two or more sets of genomic regions 
@Quinlan2010-gc. All of the aformentioned DSLs are designed to be evaluated
either at the commandline or embedded in scripts for batch processing. They 
exist in asparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analysing genomics data with R, an interactive data analysis environment that 
encouragesreproducibility and provides high-level abstractions for manipulating,
modeling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by in the aformentioned Bioconductor packages and are not easily 
extended.

The Bioconductor infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behavior that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
@Wickham2014-jc. A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. The `dplyr` pacakge [@Wickham2017-dplyr] 
defines an API that maps notions from the general relational algebra to 
operations on tidy data. It expresses each operation as a cohesive, endomorphic
verb. Taken together these features enable a user to write human
readable analysis workflows.

@@

A function is cohesive if it performs a singular task and does not produce
any side-effects. In the `plyranges` DSL our algebra for performing
genomic arithmetic is a key example of cohesion. We define anchoring operators
that decorate a GRanges object with an 'anchor' and modify the semantics
of performing genomic arithmetic. The anchoring operators amount to fixing a GRanges
object by its start, center, or end coordinates (or fixing these coordinates
by strand). This enables any arithmetic function that performs a 
coordinate transformation to remain cohesive: that is they always perfom
the same operation regardless of whether a GRanges object is anchored or not.
The only difference is the result of performing the arithimetic changes when
contextual information is added to the GRanges object.

Another example of our algebra altering object semantics, while maintaining 
cohesion is through the 'group_by' operator. Like anchoring, this operator 
decorates a GRanges object with a column name (or names) that defines a 
partitioning of the  GRanges by the unique values in the column(s). Functions 
defined in the  `plyranges` DSL that perform restriction, aggregation, or 
column modfication still perform those singular tasks, however grouping changes
how those tasks are performed.

Our algebra recasts the actions of finding overlaps or nearest neightbours 
between two genomic regions as a relational join operator. The join operator 
acts on two GRanges object, a query and a subject. The join operator is 
relational in the sense that metadata from the query and subject ranges is 
retained in the joined range.  The commonality between
all join operators in the `plyranges` DSL is to use the index of where in
the query range the subject range 'hits' the query range as a primary key 
(figure *need to make figure*). The notion of 'hits' and hence
the resulting key is determined by the type of join (preceding, nearest, follows
overlaps).

@@

comparsion to Bioconductor API here? table of join ops?

```{r olap, echo = FALSE, fig.cap="The three overlap joins: the query and subject ranges are coloured by their metadata. When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.", fig.width = 5, fig.height = 7}
a <- GRanges(seqnames = "chr1",
               strand = c("+", "-"),
               ranges = IRanges(c(1, 9), c(7, 10)),
               key.a = letters[1:2])

b <- GRanges(seqnames = "chr1",
               strand = c("-", "+"),
               ranges = IRanges(c(2, 6), c(4, 8)),
               key.b = LETTERS[1:2])

c <- join_overlap_intersect(a,b)
d <- join_overlap_inner(a, b)
e <- join_overlap_left(a, b)

p1 <- autoplot(a, aes(fill = key.a)) + 
  scale_fill_brewer(type = "qual", palette = "Set2") + 
  guides(fill = FALSE)
p2 <- autoplot(b, aes(fill = key.b))  + 
  scale_fill_brewer(type = "qual", palette = "Set1") +
  guides(fill = FALSE)
p3 <- autoplot(c, aes(fill = key.a, colour = key.b)) + 
  scale_fill_brewer(type = "qual", palette = "Set2") +
  scale_color_brewer(type = "qual", palette = "Set1") +
  guides(fill = FALSE, color = FALSE) 
p4 <- autoplot(d, aes(fill = key.a, colour = key.b)) + 
  scale_fill_brewer(type = "qual", palette = "Set2") +
  scale_color_brewer(type = "qual", palette = "Set1") +
  guides(fill = FALSE, colour = FALSE)
p5 <- autoplot(e, aes(fill = key.a, colour = key.b))  +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  scale_color_brewer(type = "qual", palette = "Set1") +
  guides(fill = FALSE, colour = FALSE)

tracks(query = p1, subject = p2, 
       `intersect` = p3, 
       `inner` = p4, 
       `left` = p5) 
```

@@

* replacing arguments with functions
* standardised inputs for reading/writing genomics data as GRanges objects

The methods `as_granges()` and `as_iranges()`
for constructing GRanges from tabular data structures, such as the _data.frame_ in
base R. These methods use non-standard evaluation so columns in a _data.frame_
can be modified before a GRanges object is created. The API also has
convenience methods for annotating or extracting annotations from _GRanges_
objects with the `set_genome_info()` and `get_genome_info()` functions.

There is a consistent framework for reading and writing 
files from and to common genomic data formats, using the _rtracklayer_ package
as a back-end @Lawrence2009-nt. The methods are implemented in the `read_/write_` family of 
functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
narrowPeaks, GFF/GTF, WIG and BigWig files.

@@

Most function calls in `plyranges` are endomorphisms: when the input 
is GRanges object the output will also be a GRanges object. The use of
endmoprhism in the `plyranges` DSL enables a user to predict the structure of 
the output of their computations and does not require them to learn any 
additional classes beyond GRanges and DataFrames.

@@

As a consequence of the design principles defined above every function in `plyranges`
performs a single action on GRanges objects. The `plyranges` DSL implments the 
core verbs from the `dplyr` package and implements a genomic relational algebra
for transforming GRanges objects (table x, y). Each verb preserves the 
semantics of GRanges object and works with derivatives of the GRanges class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a vocabulary of several verbs, rather 
than having to memorise function names that are nouns.

This approach strongly constrasts the `GenomicRanges/IRanges` OO interface, 
which emphasises the use of setter and getter methods. In that interface, 
core components and metadata are updated via replacement methods 
(requiring knowledge of the class components), while our interface requires 
only a single call to `mutate()` to perform the modification.

>>>>

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogeneous genomics data heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED.An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use an 
SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric 
Query Language (GMQL) implements a relational algebra for combining genomic datasets 
@Kaitoua2017-pw.  The command line application BEDtools develops an extensive 
algebra for performing arithmetic between two or more sets of genomic regions 
@Quinlan2010-gc. All of the aforementioned DSLs are designed to be evaluated
either at the command line or embedded in scripts for batch processing. They 
exist in sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analysing genomics data with R, an interactive data analysis environment that 
encourages reproducibility and provides high-level abstractions for manipulating,
modelling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by in the aforementioned Bioconductor packages and are not easily 
extended.

The Bioconductor infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behaviour that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
@Wickham2014-jc. A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
defines an API that maps notions from the general relational algebra to 
operations on tidy data. It expresses each operation as a cohesive, endomorphic
verb. Taken together these features enable a user to write human
readable analysis workflows.

@@

A function is cohesive if it performs a singular task and does not produce
any side-effects. In the `plyranges` DSL our algebra for performing
genomic arithmetic is a key example of cohesion. We define anchoring operators
that decorate a GRanges object with an 'anchor' and modify the semantics
of performing genomic arithmetic. The anchoring operators amount to fixing a GRanges
object by its start, center, or end coordinates (or fixing these coordinates
by strand). This enables any arithmetic function that performs a 
coordinate transformation to remain cohesive: that is they always perform
the same operation regardless of whether a GRanges object is anchored or not.
The only difference is the result of performing the arithmetic changes when
contextual information is added to the GRanges object.

Another example of our algebra altering object semantics, while maintaining 
cohesion is through the 'group_by' operator. Like anchoring, this operator 
decorates a GRanges object with a column name (or names) that defines a 
partitioning of the  GRanges by the unique values in the column(s). Functions 
defined in the  `plyranges` DSL that perform restriction, aggregation, or 
column modification still perform those singular tasks, however grouping changes
how those tasks are performed.

Our algebra recasts the actions of finding overlaps or nearest neighbours 
between two genomic regions as a relational join operator. The join operator 
acts on two GRanges object, a query and a subject. The join operator is 
relational in the sense that metadata from the query and subject ranges is 
retained in the joined range.  The commonality between
all join operators in the `plyranges` DSL is to use the index of where in
the query range the subject range 'hits' the query range as a primary key 
(figure *need to make figure*). The notion of 'hits' and hence
the resulting key is determined by the type of join (preceding, nearest, follows
overlaps).

@@

more comparison to Bioconductor API here? table of join ops?

@@

A core design principle of the `plyranges` DSL is interface consistency:
a user should not be surprised by the input or output of `plyranges` code.
A key example of consistency is how `plyranges` handles strand information.
Every function that computes with strand information indicates its intentions
by including suffixes such as 'directed', 'upstream' or 'downstream' in its
name, otherwise strand is ignored. This strongly differs from 
the Bioconductor packages, which produces surprising output by assuming the 
user is always interested in using strand in the majority of circumstances 
(unless a user is computing coverage or finding flanking ranges).

Our use of suffixes in function names, also highlights are core tenet of 
the `plyranges` design: avoid complex generic functions with many arguments
and instead use cohesive functions with a minimal number of arguments. As an
example, we have written a consistent framework for reading and writing 
files from and to common genomic data formats as GRanges, 
using the `rtracklayer` package as a back-end @Lawrence2009-nt. We have replaced
that packages generic functions for importing files with a family of reader
functions that all take the exact same arguments.

@@

Most function calls in `plyranges` are endomorphisms: when the input 
is GRanges object the output will also be a GRanges object. The use of
endomorphism in the `plyranges` DSL enables a user to predict the structure of 
the output of their computations and does not require them to learn any 
additional classes beyond GRanges and DataFrames.

@@

As a consequence of the design principles defined above every function in `plyranges`
performs a single action on GRanges objects. The `plyranges` DSL implements the 
core verbs from the `dplyr` package and implements a genomic relational algebra
for transforming GRanges objects (table x, y). Each verb preserves the 
semantics of GRanges object and works with derivatives of the GRanges class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a vocabulary of several verbs, rather 
than having to memorise function names that are nouns.

This approach strongly contrasts the `GenomicRanges/IRanges` OO interface, 
which emphasises the use of setter and getter methods. In that interface, 
core components and metadata are updated via replacement methods 
(requiring knowledge of the class components), while our interface requires 
only a single call to `mutate()` to perform the modification.
